# Standard Operating Procedure — Weekly Dashboard Update {#sop}

```{r sop-header-table, echo=FALSE}
kable(
  data.frame(
    Field = c("Document Title", "Document Number", "Version", "Effective Date",
              "Review Date", "Author", "Reviewed By", "Approved By",
              "Department", "Classification"),
    Value = c(
      "Weekly Update Procedure — ILI Surveillance Dashboard, PKK Tawau",
      "SOP-EPID-TWU-001",
      "1.0",
      format(Sys.Date(), "%d %B %Y"),
      format(Sys.Date() + 365, "%d %B %Y"),
      "[Name, Designation]",
      "[Name, Designation]",
      "[Name, Designation]",
      "Epidemiology Unit, Pejabat Kesihatan Kawasan Tawau",
      "Internal — Not for Public Distribution"
    )
  ),
  col.names = c("", ""),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("bordered"), full_width = FALSE, font_size = 13) %>%
  column_spec(1, bold = TRUE, width = "5cm") %>%
  column_spec(2, width = "10cm")
```

---

## Purpose and Scope {#sop-purpose}

This Standard Operating Procedure (SOP) describes the steps required to update the ILI Surveillance Dashboard (`ILI_Surveillance_Dashboard_v3_Week[N].html`) following the close of each epidemiological week. It is applicable to all personnel in the PKK Tawau Epidemiology Unit authorised to perform dashboard updates.

The procedure covers:

- Data collection and receipt verification
- Data extraction and validation using the Python script
- Manual update of the dashboard HTML file
- Post-update quality checks
- File versioning and distribution

This SOP does **not** cover: initial dashboard setup, threshold recalculation at the start of a new surveillance season, or modification of the dashboard's code structure. Those activities require separate technical authorisation.

---

## Roles and Responsibilities {#sop-roles}

```{r roles-table}
kable(
  data.frame(
    Role = c(
      "Data Coordinator",
      "Dashboard Operator",
      "Reviewing Officer",
      "Unit Head"
    ),
    Responsibility = c(
      "Collects facility ILI data images, age breakdown images, clinical sampling line list, and exports the notification database (inf.xlsx) each week. Verifies completeness before handover.",
      "Performs Python validation script, updates HTML arrays, conducts post-update checks, saves versioned file, and distributes to recipients.",
      "Spot-checks the updated dashboard against source data for the current week before distribution. Signs off on Section 5.9.",
      "Authorises distribution if intensity level has changed or if data anomalies are identified."
    ),
    `Designated Staff` = rep("[To be assigned]", 4)
  ),
  caption = "Roles and responsibilities for the weekly dashboard update cycle",
  col.names = c("Role", "Responsibility", "Designated Staff"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(2, width = "8cm")
```

---

## Update Schedule {#sop-schedule}

```{r schedule-table}
kable(
  data.frame(
    Day = c("Sunday (Week close)", "Monday", "Tuesday", "Wednesday", "Thursday"),
    Activity = c(
      "Epidemiological week closes. Facility staff finalise case counts.",
      "Facility data submitted to PKK Tawau. Data Coordinator collects all inputs.",
      "Data Coordinator verifies completeness of all inputs. Dashboard Operator runs Python validation.",
      "Dashboard Operator updates HTML file. Reviewing Officer performs spot-check.",
      "Finalised dashboard distributed to recipients (latest)."
    )
  ),
  caption = "Target weekly update schedule",
  col.names = c("Day", "Activity"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(2, width = "12cm")
```

> **Target:** Dashboard update should be completed and distributed no later than **Thursday** of the following week.

---

## Required Inputs {#sop-inputs}

Before beginning the update procedure, verify that **all four inputs** for the current epidemiological week (Week $N$) are available:

```{r inputs-checklist}
kable(
  data.frame(
    `#` = 1:4,
    Input = c(
      "Facility summary table image",
      "Age breakdown table image",
      "Clinical sampling line list image",
      "Notification database export (inf.xlsx)"
    ),
    Content = c(
      "KES (SEMASA), OPD (SEMASA), KONSULTASI % (SEMASA) — one row per sentinel facility, current week",
      "ILI cases and OPD by age group (< 10, 10–18, 19–59, ≥ 60) per facility, current week",
      "Sample line list rows for current week: year, week, age, gender, result columns",
      "Cumulative influenza notification line list from Week 1 to current week N"
    ),
    Format = c("Image (.png / .jpg)", "Image (.png / .jpg)", "Image (.png / .jpg)", "Excel (.xlsx)")
  ),
  caption = "Required inputs for each weekly dashboard update",
  col.names = c("#", "Input", "Content", "Format"),
  booktabs = TRUE,
  align = c("c", "l", "l", "l")
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(3, width = "7cm")
```

**Do not proceed if any input is missing.** Record the missing item in the update log and escalate to the Data Coordinator.

---

## Step-by-Step Procedure {#sop-steps}

### Step 1 — Extract Facility ILI Data (Input 1) {#step1}

From the **facility summary table image**, read and record the following values for each of the seven sentinel facilities for Week $N$:

- ILI cases ($N_{\text{ILI}}$)
- OPD attendances ($N_{\text{OPD}}$)
- Consultation rate as reported ($CR_{\text{reported}}$)

Enter values into the **Data Entry Worksheet** (Table \@ref(tab:data-worksheet)).

```{r data-worksheet, echo=FALSE}
kable(
  data.frame(
    Facility = c("KK Apas Balung", "KK Felda Umas-Umas", "KK Kalabakan",
                 "KK Merotai Besar", "KK Sungai Tamang", "KK Tawau", "KK Kinabutan",
                 "DISTRICT TOTAL"),
    Cases = rep("___", 8),
    OPD = rep("___", 8),
    `CR Reported (%)` = rep("___", 8),
    `CR Computed (%)` = rep("___", 8),
    Match = rep("☐ Yes  ☐ No", 8)
  ),
  caption = "Data entry worksheet — facility ILI data (complete for each update)",
  col.names = c("Facility", "Cases", "OPD", "CR Reported (%)", "CR Computed (%)", "Match?"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("bordered"), full_width = TRUE, font_size = 12) %>%
  row_spec(8, bold = TRUE)
```

**Compute the CR** for each facility: $CR = (Cases / OPD) \times 100$. Round to two decimal places. Compare against the reported value. A discrepancy > 0.1 percentage point requires re-reading the source image before proceeding.

**Compute the district total:** Sum all facility cases and all facility OPD counts. Compute the district CR as the pooled rate. This value becomes **OBS[N]**.

**Compute SMO[N]:** $\text{SMO}[N] = (\text{OBS}[N-2] + \text{OBS}[N-1] + \text{OBS}[N]) / 3$. Record to two decimal places.

### Step 2 — Extract Age Group Data (Input 2) {#step2}

From the **age breakdown image**, read the current-week ILI case counts for each of the four age groups (< 10, 10–18, 19–59, ≥ 60) for each facility. These values replace the `AGE_CUR_CASES` array.

For `AGE_CUM_CASES`, add the current week's values to the previous cumulative totals for each facility and age group. Verify that the row totals equal the facility's total case count from Step 1.

### Step 3 — Run Python Validation Script (Input 4) {#step3}

Before reading the clinical sampling or notification data, run the Python validation script against the current `inf.xlsx` file:

```bash
python3 validate_notifications.py --file inf.xlsx --week N
```

The script outputs:
- Per-week counts for Influenza A, Influenza B, and Unspecified/Other
- A MATCH / MISMATCH flag comparing weeks 1 to $N-1$ against the current dashboard arrays
- Week $N$ new values for the NOTIF_INFA, NOTIF_INFB, NOTIF_OTHER arrays

**All prior weeks must show MATCH before proceeding.** If a MISMATCH is found in any prior week, investigate and resolve before updating the current week.

Record the Week $N$ notification values:

- **NOTIF_INFA[N]:** `___`
- **NOTIF_INFB[N]:** `___`
- **NOTIF_OTHER[N]:** `___`
- **Total notifications Week N:** `___`

Also extract age group data for Week $N$ notifications from the script output for the `NOTIF_AGE_BY_WK` object, and update the `NOTIF_AGE_CUMUL` cumulative totals.

### Step 4 — Extract Sampling Data (Input 3) {#step4}

From the **clinical sampling line list image**, count the following for Week $N$:

| Value | Count | Notes |
|---|---|---|
| Total samples tested | ___ | All rows for Week N |
| Negative | ___ | No pathogen detected |
| Influenza A positive | ___ | PATH_BY_WEEK[N][0] |
| Influenza B positive | ___ | PATH_BY_WEEK[N][1] |
| SARS-CoV-2 positive | ___ | PATH_BY_WEEK[N][2] |
| Untyped positive | ___ | PATH_BY_WEEK[N][3] |
| Inconclusive | ___ | PATH_BY_WEEK[N][4] |
| Total positive | ___ | Sum of all positives |

**Verify:** Total samples = Negative + Total positive + Inconclusive.

Compute positivity rate: $\text{Yield} = (\text{Total positive} / \text{Total samples}) \times 100$ (round to one decimal place).

Determine the `remarks` string for the SAMPLE_DATA entry based on pathogens detected.

Record age and gender breakdown from the line list for the `SAMPLE_CHARS[N]` object.

### Step 5 — Open and Edit the Dashboard HTML File {#step5}

Open the **most recent versioned dashboard file** (e.g. `ILI_Surveillance_Dashboard_v3_Week[N-1].html`) in a text editor. **Save immediately as a new file:**

```
ILI_Surveillance_Dashboard_v3_Week[N].html
```

Make **no edits** to the prior week's file. All changes are made only to the new file.

### Step 6 — Update JavaScript Data Arrays {#step6}

Using the values recorded in Steps 1–4, update the following arrays and values in the `<script>` section of the HTML file. Each item should be updated in sequence.

```{r array-checklist}
kable(
  data.frame(
    `#` = 1:14,
    Array = c("OBS", "SMO", "NOTIF overlay in mainChart",
              "FACILITIES[0] — KK Apas Balung", "FACILITIES[1] — KK Felda U-U",
              "FACILITIES[2] — KK Kalabakan", "FACILITIES[3] — KK Merotai Besar",
              "FACILITIES[4] — KK Sungai Tamang", "FACILITIES[5] — KK Tawau",
              "FACILITIES[6] — KK Kinabutan",
              "FAC_WEEKS", "AGE_CUR_CASES", "AGE_CUM_CASES",
              "SAMPLE_DATA"),
    `Values to append` = c(
      "OBS[N] — district CR for Week N",
      "SMO[N] — 3-week trailing average",
      "Notification total for Week N in the bar dataset",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "cases[N], opd[N], rate[N], updated total_c, total_o",
      "Append N to the array",
      "Replace all 7 facility rows with current week values",
      "Replace all 7 facility rows with cumulative values",
      "Append new {wk:N, samples:_, neg:_, pos:_, inc:_, yield:_, remarks:'_'}"
    ),
    Done = rep("☐", 14)
  ),
  caption = "Array update checklist — complete in sequence for each weekly update",
  col.names = c("#", "Array / Variable", "Values to Update", "Done"),
  booktabs = TRUE,
  align = c("c", "l", "l", "c")
) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = TRUE, font_size = 12) %>%
  column_spec(3, width = "7cm")
```

```{r array-checklist-2}
kable(
  data.frame(
    `#` = 15:22,
    Array = c("PATH_BY_WEEK", "SAMPLE_CHARS", "NOTIF_WEEKS", "NOTIF_INFA",
              "NOTIF_INFB", "NOTIF_OTHER", "NOTIF_META", "NOTIF_AGE_BY_WK / CUMUL"),
    `Values to append` = c(
      "Append [InfA, InfB, CoV2, Untyped, Inc] for Week N",
      "Append N: {u18:_, o18:_, m:_, f:_}",
      "Append N",
      "Append Influenza A count for Week N",
      "Append Influenza B count for Week N",
      "Append Unspecified/Other count for Week N",
      "Append {u18:_, o18:_} for Week N",
      "Add entry for Week N; update CUMUL totals"
    ),
    Done = rep("☐", 8)
  ),
  col.names = c("#", "Array / Variable", "Values to Update", "Done"),
  booktabs = TRUE,
  align = c("c", "l", "l", "c")
) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = TRUE, font_size = 12) %>%
  column_spec(3, width = "7cm")
```

### Step 7 — Update Static Text Labels {#step7}

Search the file for all occurrences of `Weeks 1–[N-1]` and replace with `Weeks 1–[N]`. These appear in:

- Weekly Values bar chart description
- Facility Activity chart description
- Weekly Sampling Log description
- Notifications chart description
- Age comment in code

Also update the notification metric card values (total notifications, subtype counts and percentages, dominant week ranges).

Also add **Week N** as a new `<option>` entry in the notification age week selector dropdown.

### Step 8 — Check and Adjust Chart Y-Axis Maxima {#step8}

If the current week's values approach or exceed prior axis maxima, increase the `max` property of the relevant chart scale:

| Chart | Scale | When to increase |
|---|---|---|
| Main chart | `y.max` (default 28) | If OBS[N] or any threshold approaches 28 |
| Bar chart | `y.max` (default 20) | If OBS[N] > 19 |
| Notifications overlay | `yNotif.max` | If any weekly total > (max − 3) |
| Pathogen chart | `y.max` | If PATH_BY_WEEK stacked total for any week > max |
| Sample positivity | `yRate.max` | If any weekly yield > (max − 3) |

### Step 9 — Post-Update Validation {#step9}

After completing all edits, open the updated HTML file in a web browser and verify:

```{r post-update-checklist}
kable(
  data.frame(
    Check = c(
      "Header data currency badge shows correct week number and date",
      "Sidebar: observed CR matches Step 1 district total",
      "Sidebar: smoothed CR matches Step 1 computed SMO[N]",
      "Sidebar: intensity status tag colour and label are correct",
      "Main chart: 2026 line extends to Week N with correct endpoint",
      "Bar chart: Week N bar present with correct colour coding",
      "Facility cards: all 7 show Week N rate and correct directional arrow",
      "Heatmap: Week N column present for all facilities",
      "Sampling metric cards: totals are cumulative through Week N",
      "Notification metric cards: totals match inf.xlsx line list",
      "No JavaScript console errors (press F12 in browser)"
    ),
    Result = rep("☐ Pass  ☐ Fail", 11)
  ),
  caption = "Post-update validation checklist",
  col.names = c("Check", "Result"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = TRUE, font_size = 12) %>%
  column_spec(2, width = "4cm")
```

Any failed check must be investigated and resolved before proceeding to Step 10.

### Step 10 — Reviewing Officer Sign-Off {#step10}

The Reviewing Officer independently verifies the following against source data:

- District CR for Week N (from Step 1 worksheet)
- Notification total for Week N (from Step 3 script output)
- Intensity classification displayed in the status tag

Upon satisfactory review, the Reviewing Officer signs Section \@ref(signoff) of this record.

### Step 11 — Save, Version, and Distribute {#step11}

1. Save the final HTML file as `ILI_Surveillance_Dashboard_v3_Week[N].html`
2. Archive the previous week's file — do not delete
3. Record the update in the Version Control Log (Section \@ref(version-log))
4. Distribute to recipients per the current distribution list

---

## Known Data Anomalies {#sop-anomalies}

The following pre-existing anomalies are documented and should be carried forward without correction unless explicitly reviewed and resolved:

```{r anomalies-table}
kable(
  data.frame(
    Anomaly = c("KK Kinabutan age-group reconciliation discrepancy"),
    Description = c(
      "The sum of age-group case counts for KK Kinabutan does not equal the facility's total case count by 1 case. This discrepancy originates in the source data prior to Week 22, 2026 and has been carried forward intentionally."
    ),
    `First noted` = c("Week 22, 2026"),
    Status = c("Known — carry forward")
  ),
  caption = "Pre-existing data anomalies — carry forward without correction",
  col.names = c("Anomaly", "Description", "First Noted", "Status"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = TRUE, font_size = 12) %>%
  column_spec(2, width = "8cm")
```

---

## Reviewing Officer Sign-Off {#signoff}

```{r signoff-block}
kable(
  data.frame(
    Field = c("Epidemiological Week Updated", "Date of Update",
              "Dashboard Operator (Name)", "Dashboard Operator (Signature)",
              "Reviewing Officer (Name)", "Reviewing Officer (Signature)",
              "Intensity Level (current week)", "Remarks"),
    Entry = rep("", 8)
  ),
  col.names = c("Field", "Entry"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("bordered"), full_width = FALSE, font_size = 13) %>%
  column_spec(1, bold = TRUE, width = "6cm") %>%
  column_spec(2, width = "8cm")
```

---

## Version Control Log {#version-log}

```{r version-log}
kable(
  data.frame(
    Version = c("SOP-EPID-TWU-001 v1.0", rep("", 4)),
    Date = c(format(Sys.Date(), "%d %b %Y"), rep("", 4)),
    `Weeks Covered` = c("1–23", rep("", 4)),
    `Changed By` = c("[Name]", rep("", 4)),
    `Summary of Changes` = c("Initial SOP release", rep("", 4))
  ),
  caption = "SOP version control log",
  col.names = c("Version", "Date", "Weeks Covered", "Changed By", "Summary of Changes"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("bordered"), full_width = TRUE, font_size = 12) %>%
  column_spec(5, width = "6cm")
```
