# Technical Reference {#technical}

---

## Technology Stack {#tech-stack}

The ILI Surveillance Dashboard is implemented as a **single self-contained HTML file** with no server-side components or database dependencies. All data are embedded directly in the file as JavaScript arrays.

```{r tech-stack-table}
kable(
  data.frame(
    Component = c("Core structure", "Styling", "Charting", "Typography",
                  "Data validation tooling", "Version control"),
    Technology = c(
      "HTML5 / CSS3 / Vanilla JavaScript (ES6+)",
      "Custom CSS with CSS custom properties (design tokens); responsive via media queries",
      "Chart.js v4.4.1 (CDN: cdnjs.cloudflare.com)",
      "DM Sans, DM Mono (Google Fonts CDN)",
      "Python 3 with openpyxl / pandas for xlsx parsing and pre-update validation",
      "Manual file versioning (ILI_Surveillance_Dashboard_v3_Week[N].html)"
    )
  ),
  caption = "Dashboard technology stack",
  col.names = c("Component", "Technology / Implementation"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) %>%
  column_spec(2, width = "10cm")
```

---

## File Architecture {#tech-file}

The dashboard is a single `.html` file containing:

- **HTML structure** — semantic layout (header, sidebar, main) with responsive grid
- **Embedded CSS** — all styling in a `<style>` block within `<head>`; includes light/dark theme tokens, responsive breakpoints, and chart-specific styles
- **Embedded JavaScript** — a single `<script>` block at end of `<body>` containing:
  - All data arrays (historical curves, observed data, facility data, sampling data, notification data)
  - Chart.js instantiation for all 8+ charts
  - Theme toggle, tab switching, and toggle button handlers
  - Summary computation function (`computeAllSummaries()`) called on page load

### Data Array Reference

```{r array-ref-table}
kable(
  data.frame(
    Array = c("AVG", "CI_UP", "OBS", "SMO",
              "OBS_2022 / OBS_2023 / OBS_2024 / OBS_2025",
              "THR", "PEAKS", "FACILITIES[i].cases/.opd/.rate",
              "FACILITIES[i].total_c/.total_o",
              "FAC_WEEKS", "AGE_CUR_CASES", "AGE_CUM_CASES",
              "SAMPLE_DATA", "PATH_BY_WEEK", "SAMPLE_CHARS",
              "NOTIF_INFA/B/OTHER", "NOTIF_META", "NOTIF_AGE_BY_WK",
              "NOTIF_AGE_CUMUL"),
    Length = c("52", "52", "52 (nulls for future weeks)", "52 (nulls)",
               "52 each", "Object (4 keys)", "4 entries",
               "N weeks each", "Scalar (cumulative)",
               "N weeks", "7×4 matrix", "7×4 matrix",
               "N entries (objects)", "N×5 matrix",
               "Object keyed by week", "N each", "N entries",
               "Object keyed by week", "6-element array"),
    Updated = c("Season start only", "Season start only",
                "Weekly — append OBS[N]", "Weekly — append SMO[N]",
                "Season start only", "Season start only",
                "Season start only",
                "Weekly — append week N value",
                "Weekly — recompute from cumulative",
                "Weekly — append N",
                "Weekly — replace with current week values",
                "Weekly — replace with cumulative totals",
                "Weekly — append new object",
                "Weekly — append new row",
                "Weekly — append new entry",
                "Weekly — append count for week N",
                "Weekly — append {u18, o18} for week N",
                "Weekly — add entry for week N",
                "Weekly — recompute from cumulative")
  ),
  caption = "JavaScript data array reference guide",
  col.names = c("Array / Variable", "Length / Structure", "Update Frequency"),
  booktabs = TRUE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11) %>%
  column_spec(3, width = "5cm")
```

---

## Browser Compatibility {#tech-browser}

The dashboard has been tested in and is supported on:

| Browser | Minimum version | Notes |
|---|---|---|
| Google Chrome | 90+ | Recommended |
| Microsoft Edge | 90+ | Full support |
| Mozilla Firefox | 88+ | Full support |
| Safari (macOS) | 14+ | Full support |
| Safari (iOS) | 14+ | Mobile responsive; touch tooltips available |
| Internet Explorer | Any | **Not supported** |

The dashboard uses CSS custom properties (variables), CSS Grid, and ES6 JavaScript features. Internet Explorer is incompatible with all three.

**CDN dependency:** The dashboard requires an active internet connection to load Chart.js from `cdnjs.cloudflare.com` and fonts from `fonts.googleapis.com`. For use in environments without internet access, these assets must be bundled locally (see Section \@ref(tech-offline)).

---

## Offline / Air-Gapped Use {#tech-offline}

To make the dashboard fully self-contained for offline use:

1. Download Chart.js: `https://cdnjs.cloudflare.com/ajax/libs/Chart.js/4.4.1/chart.umd.js`
2. Save to the same directory as the dashboard HTML file
3. Replace the CDN `<script>` tag in the HTML `<head>`:

```html
<!-- Replace this: -->
<script src="https://cdnjs.cloudflare.com/.../chart.umd.js"></script>

<!-- With this: -->
<script src="chart.umd.js"></script>
```

4. For Google Fonts offline: either embed the font files locally or remove the `<link>` tag (the dashboard will fall back to system sans-serif fonts).

---

## File Naming Convention {#tech-naming}

```
ILI_Surveillance_Dashboard_v3_Week[N].html
```

| Segment | Meaning |
|---|---|
| `ILI_Surveillance_Dashboard` | System name |
| `v3` | Major template version (structural/design changes increment this) |
| `Week[N]` | Most recent epidemiological week for which data are entered |

Prior week files should be **retained and not deleted**. They form an audit trail of the surveillance record.

---

## Python Validation Script {#tech-python}

The pre-update validation script (`validate_notifications.py`) performs the following checks against the `inf.xlsx` export:

1. Reads all rows for year 2026 from `Sheet1`
2. Applies the subtype classification rule (see Section \@ref(notif-method))
3. Outputs per-week counts for Influenza A, Influenza B, and Unspecified/Other from Week 1 to the most recent week in the file
4. Compares weeks 1 to $N-1$ against the hardcoded expected values from the current dashboard arrays
5. Flags any MISMATCH and exits with a non-zero return code if any prior week does not match

The script requires Python 3.7+ and the `openpyxl` package:

```bash
pip install openpyxl
python3 validate_notifications.py --file /path/to/inf.xlsx --week 23
```

---

## Suggested Future Enhancements {#tech-future}

The following enhancements are identified for potential future development:

| Enhancement | Description | Priority |
|---|---|---|
| Automated data ingestion | Python or R script to parse source images (OCR) and auto-populate arrays, reducing manual transcription errors | High |
| PDF export | Add a print-optimised CSS stylesheet or a button to export the dashboard as a PDF report | Medium |
| Season recalculation workflow | Document or automate the WHO ACM upload and threshold extraction at the start of each new surveillance year | High |
| Historical archive page | Embedded archive of prior-season dashboard files for longitudinal review | Low |
| Alert notification | Email or messaging alert triggered when intensity classification changes | Medium |
| Mobile push | Progressive Web App (PWA) wrapper to enable dashboard access and push notifications on mobile devices | Low |

---

## References {#references}

```{r, results='asis', echo=FALSE}
cat("<!-- References are managed in references.bib and rendered automatically by bookdown -->")
```
